# (PART) Regression and Linear Models {-}

# Correlation and Regression {#correg}

```{r, echo=FALSE}
# clear out workspace
#rm(list=ls())
```

## Introduction {#introcorreg}

We can think about ANOVA in terms of explaining a continuous variable by the different groups, a nominal variable, but what if both variables are continuous? In this case, we might still be interested in ascertaining whether there is a relationship between the variables and correlation can help. If we want to try and explain one variable with the other variable, we fit a simple linear regression model and this allows us to describe the relationship with an equation.

In this chapter we look at the two basic statistical concepts of correlation and simple linear regression.

## Correlation

To illustrate the concept of correlation, one might ask the question "Is left leg length (LLL) associated with total height of a person?" 

```{r scatplot1, echo=F, eval=T, fig.cap='Scatterplot of total height and left leg length.'}
set.seed (101)
LLL <- rnorm (100,110,4)
LRL <- rnorm(100, LLL,0.05)
TotalHeight <- (LLL+LRL)/2*1.8/1.1 +rnorm (100,0,4)

plot (LLL, TotalHeight, xlab="LLL (cm)", ylab="Total Height (cm)", pch=20)

# save data
hgt <- data.frame(LLL=LLL, LRL=LRL, TotalHeight=TotalHeight)
```

Does Figure \@ref(fig:scatplot1) provide evidence that there is an association or __correlation__, as we say with continuous variables?

If you think that plot provided evidence of correlation, what about this data set (Figure \@ref(fig:scatplot2))? 

```{r scatplot2, echo=F, eval=T,fig.cap='Scatterplot of total height and left leg length.' }
set.seed (102)
LLL <- rnorm (100,110,4)
LRL <- rnorm(100, LLL,0.05)
TotalHeight <- (LLL+LRL)/2*1.8/1.1 +rnorm (100,0,8)

plot (LLL, TotalHeight, xlab="LLL (cm)", ylab="Total Height (cm)", pch=20)

```

Or this one Figure \@ref(fig:scatplot3)? 

```{r scatplot3, echo=F, eval=T, fig.cap='Scatterplot of total height and left leg length.'}
set.seed (103)
LLL <- rnorm (100,110,4)
LRL <- rnorm(100, LLL,0.05)
TotalHeight <- (LLL+LRL)/2*1.8/1.1 +rnorm (100,0,40)

plot (LLL, TotalHeight, xlab="LLL (cm)", ylab="Total Height (cm)", pch=20)

```

It would be useful to have a method to objectively answer these questions. In this case, one undertakes correlation. Correlation asks the question "is there a relationship?" but not what the relationship is - for that we use regression (which we consider later). Here we will consider linear correlation but there could also be a non-linear relationship and then the statistic of association is called __concurvity__. 

Correlation is typically indexed by a __correlation coefficient__ ($R$ or $r$) which takes a value from -1 to +1 where, 

+ -1 indicates a perfect negative relationship, 
+ 0 means no relationship and 
+ +1 indicates a perfect positive relationship.   

The statistic $r$ is the estimated (or sample) coefficient of the unknown population correlation coefficient, $\rho$. 

There are many types of correlation coefficients, but one frequently used is Pearson's product moment correlation coefficient which, for two continuous variables denoted by $x$ and $y$, is given by:  

$$r=\frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2\sum_{i=1}^{n}(y_i-\bar{y})^2}}$$
where 

+ $n$ is the number of observations,
+ $x_i$ and $y_i$ are values for observation $i$,
+ $\bar x$ and $\bar y$ are sample means for each variable.

For example, consider the contribution of left leg length to human height in a sample of men (using the data in Figure \@ref(fig:scatplot1)), we can use R to do the calculation; the variables `LLL` and `TotalHeight` are stored in a data frame `hgt`. 

```{r , echo=F, eval=F, fig.cap = 'Scatterplot of total height and left leg length.'}
set.seed (101)
LLL <- rnorm (100,110,4)
LRL <- rnorm(100, LLL,0.05)
TotalHeight <- (LLL+LRL)/2*1.8/1.1 +rnorm (100,0,4)


plot (LLL, TotalHeight, xlab="LLL (cm)", ylab="Total Height (cm)", pch=20)

```
$$r=\frac{\sum(diffx)(diffy)}{\sqrt{\sum (diffx)^2\sum(diffy)^2}}$$
where $diffx = x_i - \bar x$ and $diffy = y_i - \bar y$. 

```{r, echo=T, eval=T}
# Manual calculation of r
diffx <- hgt$LLL - mean(hgt$LLL) ####differences of x and xbar
diffy <-  hgt$TotalHeight - mean(hgt$TotalHeight)

r <- sum(diffx*diffy)/sqrt(sum (diffx^2)*sum(diffy^2) )
print (r)
```

Thus, $r = 0.868$; it is close to $+1$ indicating a strong, positive relationship between LLL and total height. 

The value can also be obtained by using the command `cor` where you supply it the two vectors of interest. 

```{r, echo=T, eval=T}
cor(hgt$TotalHeight, hgt$LLL)
```

Another alternative  formula for the sample product-moment correlation coefficient is

$$r= \frac{\frac{\sum (x-\bar{x})(y-\bar{y})}{n-1}}{s_xs_y}$$

Or assuming the population correlation coefficient is of interest
$$\rho= \frac{\frac{\sum (x-\bar{x})(y-\bar{y}}{N}}{\sigma_x\sigma_y}$$


The numerator here is called the __covariance__ of $x$ and $y$ and so an alternative way to describe the formula is: 

$$r = \frac{Cov(xy)}{\sqrt{Var(x)Var(y)}}$$ 

***

**Q13.1** I have two random variables $X$ and $Y$. The variances of these variables are 2.5 and 4, respectively, and their covariance is -2.5. What is the correlation between these two variables?

**Q13.2** Suppose I was to create a new random variable $Z=X + Y$ from the previous question. What would be the consequence to the variance estimate of $Z$ if I were to ignore the covariance of $X$ and $Y$ (assuming X and Y actually were correlated)?  

### Significance of $r$

Just like other statistics $r$ can have a significance associated with it. The test is typically whether $\rho$, the unknown population correlation coefficient, is different from one. So 

$$H_0:  \rho = 0$$
$$H_1:  \rho \neq 0$$

In fact, the significance is generated by a $t$ test statistic with $n-2$ degrees of freedom:

$$t = \frac{r \times \sqrt{n-2}}{\sqrt {1-r^2}}$$

In the case of the correlation of total height and left leg length, $r = 0.868$ and $n=100$ so 

$$t = \frac{0.868 \times \sqrt{100-2}}{\sqrt {1-0.868^2}}= \frac{8.593}{0.497} = 17.289$$

The significance level associated with this test statistic is found from:

```{r, echo=T, eval=T}
2*pt(q=17.289, df=98, lower.tail=FALSE) ###assuming a two-tailed test 
```

Such a small $p$-value is perhaps not surprising in this case where $r$ is close to one. 

The confidence intervals for $\rho$, the unknown population correlation coefficient are actually quite complicated involving a transformation of $r$ to normalise it,  then adding/subtracting an equivalent of the "$se \times t_{\alpha/2, df}$" term seen so frequently throughout this module, and then back-transforming back to scale of $r$. 

Unsurprisingly there is a function in R to compute the CI and conduct the hypothesis test. 
```{r, echo=T, eval=T}
cor.test (hgt$TotalHeight, hgt$LLL)
```

### Correlation and causation

The causal correlation fallacy is the idea that just because there is a *correlation*, or indeed *association*, (in the case of categorical variables) between two (or more) sets of variables then there is a causal link. Obviously, causality implies correlation but correlation does not necessarily imply causation. 

Variables might be correlated by

a) chance
b) another third variable which affects them both. 
c) genuine causation.
 
Figure \@ref(fig:quartet) shows "Anscombe's quartet", a series of famous data sets that show identical correlation coefficients but probably negligible causation! The summary values of each variable are shown below. 

```{r, echo=F, eval=T}
require(stats); require(graphics)
#summary(anscombe)
with(cor.test (x1, y1), data = anscombe)
with(cor.test (x2, y2), data = anscombe)
with(cor.test (x3, y3), data = anscombe)
with(cor.test (x4, y4), data = anscombe)
```

Each pair of variables (i.e.$x_1$ and $y_1$, $x_2$ and $y_2$, etc.) have identical $r$ values!  

```{r quartet, echo=F, eval=T, fig.cap = "Anscombe's four regressions"}
##-- now some "magic" to do the 4 regressions in a loop:
ff <- y ~ x
mods <- setNames(as.list(1:4), paste0("lm", 1:4))
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  ## or   ff[[2]] <- as.name(paste0("y", i))
  ##      ff[[3]] <- as.name(paste0("x", i))
  mods[[i]] <- lmi <- lm(ff, data = anscombe)


# print(anova(lmi))
}

## See how close they are (numerically!)
#sapply(mods, coef)
#lapply(mods, function(fm) coef(summary(fm)))

## Now, do what you should have done in the first place: PLOTS
op <- par(mfrow = c(2, 2), mar = 0.1+c(4,4,1,1), oma =  c(0, 0, 2, 0))
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  plot(ff, data = anscombe, col = "red", pch = 21, bg = "orange", cex = 1.2,
       xlim = c(3, 19), ylim = c(3, 13))
  abline(mods[[i]], col = "blue")
}
par(op)
```

Here is a slightly more modern version from the R library `datasauRus`. All the following data have an approximately identical albeit low $r$. 

```{r datasaurus, echo=F, fig.cap = "Datasaurus regressions"}

library(ggplot2)
library(datasauRus)

par (mfrow=c(3,4))


names1 <- unique (datasaurus_dozen$dataset)
for (i in 1:12){
  temp <- datasaurus_dozen[datasaurus_dozen$dataset==names1[i],]

temp<- as.data.frame (temp)
x<- temp$x
y <- temp$y

 print (names1[i])
 print ( cor.test (x,y))
 plot (x,y, main=names1[i], pch=20)
}

```

**Q13.3** A television advert for _Booster_ breakfast cereal claims "that people who start their day with a healthy breakfast like _Booster_ actually lose more weight than those who skip breakfast."

Does this support a causal link between healthy breakfast cereals and weight loss? 

\newpage

## Regression

Correlation asks the question "Is there a (linear) relationship?" A more interesting question might be "Is there a (linear) relationship and what is it?" i.e. what is our best estimate of the equation relating the two variables; __linear regression__ allows us to do this. Here we will explore this using the environmental impact assessment (EIA) data and other datasets we have already encountered, for example, we might use variables in the EIA data to predict depth (`depth`) at a particular location.

Regression is a way to study relationships between variables. There are three main reasons why we may want to do this:

+ __Description__: It can be useful to describe relationships (without necessarily really explaining them. For example a spatial map of an animal species for example. 
+ __Explanation:__ Genuine interest in the nature of the relationship between variables e.g. How is depth and penguin density related?
+  __Prediction:__ Using variables to predict others (e.g. using  `DistCoast` to predict `depth`)

Linear regression models: 

+ contain explanatory (sometime called "independent") variable(s) which help us explain or predict the behaviour of the response variable.
+ assume constantly increasing, or decreasing, relationships between each explanatory variable and the response. 

In simple linear regression, we consider only one explanatory variable in the regression model. 

### Exploratory data analysis

To analyse the EIA data properly would require some more advanced methods, but we can illustrate the basic principles of simple linear regression with these data. 

```{r, echo=F, eval=T, warning=F, message=F}
  #load libraries and data
  library(ggplot2, quietly=T)
 # library(tidyr, quietly=T)
  library(dplyr, quietly=T)
  library (gridExtra, quietly=T)
  EIAData<- read.csv("data/EIA_dataset_for_lectures_V1.0.csv", header=T)
  
  # filter out near-shore data for our current purposes
  workingData<- EIAData%>%filter(DistCoast>100)

  options(digits=7)
```

First, we consider a potential relationship between the distance from the coast and the depth of the water (fairly trivial but it will illustrate the methods):

+ We want to use a function of distance from coast to explain depth.
+ Visualising the relationship between two numeric (and continuous) variables suggests using a scatterplot.
+ By convention, we put distance from coast on the $x$-axis because this is the explanatory variable (and the function we are after is $y=f(x)$).

```{r, echo=FALSE, eval=TRUE,fig.cap='Relationship of distance to coast and depth.'}
p<- ggplot(x=DistCoast, y=depth, data=workingData) +
  geom_point(aes(DistCoast, depth), color='blue', alpha=0.3)+
  xlab('Distance from coast (m)')+
  ylab('Depth (m)')

```

```{r depthdistscatter, echo=FALSE, eval=TRUE, fig.cap='Relationship of distance to coast and depth.', dev='png'}
p
```

The scatterplot (Figure \@ref(fig:depthdistscatter)) tells us:

+  waters nearer the shore are shallower
+  there is a  positive relationship is apparent (i.e. as distance to coast increases so does the depth)
+ (there are also striations, which is interesting)

How can we formalise this relationship?

### Model specification

#### Setting up the model

We will first explain this using generic data. Assume there is a variable $X$ and a variable $Y$,
which is thought to be potentially dependent on $X$. We can plot them out (Figure \@ref(fig:xy)). 

```{r xy, echo=FALSE, eval=TRUE, fig.cap="A scatterplot illustrating the relationship between $X$ and $Y$.", dev='png'}
set.seed(101)
x <-seq (1,100,2)
y <- 10+ 2*x+rnorm (50,0,20)

model1 <- lm (y~x)

par (xaxs="i")
par (yaxs="i")

plot (x,y, pch=20, xlab="X", ylab="Y")
```

\newpage 

A linear/straight line relationship between $X$ and $Y$ might be a reasonable starting point; perhaps some lines like those in  Figure \@ref(fig:manylines) would be a good fit and summarise these data well. 

```{r manylines, echo=FALSE, eval=TRUE, fig.cap="Scatterplot with examples of possible fitted lines."}

set.seed(101)
x <-seq (1,100,2)
y <- 10+ 2*x+rnorm (50,0,20)

par (xaxs="i")
par (yaxs="i")
plot (x,y, pch=20, xlab="X", ylab="Y")

abline(a=-5, b=2.5, col=2, lty=2, lwd=2)
abline(a=5, b=2, col=4, lwd=2)
abline(a=15, b=1.9, col=3, lty=4, lwd=2)

```
All the lines shown in Figure \@ref(fig:manylines) have the same general form. What we want to do is find the 'best' model. __A simple linear regression model__ has the form:

$$\textrm{response} = \textrm{intercept} + \textrm{slope} \times \textrm{explanatory variable} + \textrm{error}$$
In notation form this can be represented as:

$$y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$

where

+ $y_i$ refers to the individual values of $Y$ indexed by $i$,i.e. i = 1, ..., n observations. This the response or the dependent variable. 
+ $x_i$ refers to the individual values of $X$ indexed by $i$,  
+ $\beta_0$ is the intercept parameter, 
+ $\beta_1$ is the slope parameter, and
+ $\epsilon_i$ is an error term.
<br/>
We use the data to estimate values for the intercept and slope. 

#### The intercept ($\beta_0$)

The intercept can be thought of in a few ways:

+  The response value (under the model) when the explanatory variable(s) is/are zero
+  Where the regression line cuts the vertical axis
+  The expected value of the response ($y_i$) when $x_i=0$.

#### The slope ($\beta_1$)

The slope, or __gradient__, of the regression line is:

+  the expected change in the response ($y_i$) when $x_i$ increases by 1 unit.

#### The error term (a model for the noise)

A linear regression model might summarise the relationship between $X$ and $Y$, but not all the observations follow this linear relationship __exactly__. The error term ($\epsilon_i$) allows for deviations from this linear relationship:

+ In the simplest version of regression described here (i.e. one explanatory variable) the error is assumed to be distributed normally in the $y$ dimension i.e. the uncertainty is in the dependent variable not the x variable.
+  The normal distribution has two parameters that describe it, the mean ($\mu$) and variance ($\sigma^2$).
+  Since we are modelling the mean response, there is __zero__ mean difference\footnote{nothing left over on average} between the line and the observations; 
+  The variance of the errors  ($\sigma_e^2$) is estimated as a part of the linear model fitting process.

This can be summarised as $\epsilon_i \sim N(0, \sigma_e^2)$. 

### Which straight line to choose?

There are many possible straight lines as we saw in Figure \@ref(fig:manylines):

+ We want values of $\beta_0$ and $\beta_1$ that look most plausible in light of our data

+ We want $\beta_0$ and $\beta_1$ which give the best fitting line - the __regression line__.

We can use least-squares to find the best fitting regression line.


#### The Least Squares (LS) criterion

A variety of criteria could be used to fit a "best fit" line. One often used criterion is the leat squares criterion. We want to choose values for the parameters that minimise the sum of the squared differences between the observed data ($y_i$) and the predictions under the model ($\hat{y}_i$). The LS criterion finds parameter estimates which minimise this: 

$$\sum_{i=1}^n (data-model)^2=\sum_{i=1}^n (y_i-\hat{y}_i)^2 = SS_{Res} $$

+ The solid line in Figure \@ref(fig:bestline) (our model for the signal) will be as close as we can get to the data (on average, based on vertical distances). Other fitted lines (e.g. like the red dashed line) will have a far higher sum of squared differences $SS_{Resl}$.

+ Note other (possibly non linear) models may be better, but this is our best straight line model.

+ The vertical distances between the observed data and the best fit line are called __"residuals"__.  The least square criterion obtains a line that minimises the summed squares of the residuals, typically abbreviated to __"the sum of squares"__.  

```{r bestline, echo=FALSE, eval=TRUE, fig.cap="A simple *X-Y* scatterplot with different regression lines; best fit line (black), less optimal line (red). The vertical difference (shown by the arrows) between the observations and the best fit line are the residuals."}
set.seed(101)
x <-seq (1,100,2)
y <- 10+ 2*x+rnorm (50,0,20)

par (xaxs="i")
par (yaxs="i")
plot (x,y, pch=20,xlab="X", ylab="Y")

abline(model1)
arrows(x[34], y[34], x1 = x[34], y1 = fitted (model1)[34], length = 0.1, angle = 30,
       code = 2, col = 1, lty = 1,
       lwd = 1)
arrows(x[25], y[25], x1 = x[25], y1 = fitted (model1)[25], length = 0.1, angle = 30,
       code = 2, col = 1, lty = 1,
       lwd = 1)
      
abline (a=20, b=1,lty=2, col=2)

```

This [shiny app](https://tullimonstrum.shinyapps.io/shinyapps/ ) allows you to choose a best fit line yourself. Can you find the best fit that minimises the sum of the square of the residuals? The red arrows indicate the residual lengths.


### Fitting the model: the details

The slope and intercept estimates can be found from the data using:

$$\hat{\beta}_1=\frac{\sum_{i=1}^{n}(x_i-\bar{x})y_i}{\sum_{i=1}^{n}(x_i-\bar{x})^2}$$

$$\hat{\beta}_0=\bar{y} - \hat{\beta}_1\bar{x}$$
where $\bar{x}$ is the mean of the explanatory variable and $\bar{y}$ is the mean of the response.

Least squares is a useful criterion and it has another advantage; the least squares estimate for the gradient is also *the maximum likelihood estimator* for the gradient which has theoretical usefulness in more advanced applications. 

### Predictions

Having obtained estimates for $\beta_0$ and $\beta_1$, they can be used to obtain predicted, or fitted, values of the response:

$$\hat{y_i}=\hat{\beta}_0+\hat{\beta}_1x_{i}$$

We can then estimate $Y$ for any given value of $X$ (within reason).

### The variance estimate

We can find the variance estimate for the error term ($\sigma_e^2$) as follows:
\begin{equation}
s^2 = \hat{\sigma_e}^2=\frac{1}{n-k-1}\sum_{i=1}^n(y_i-\hat{y}_i)^2
\end{equation}

where 

+ $\hat{y}_i$ are the fitted values, 
+ $n$ is the number of observations, 
+ $k$ is the number of slope parameters estimated (in simple linear regression $k=1$). 

This estimate ($s=\sqrt{s^2}$) is provided as the `Residual Standard Error` in the `R` output (see later):

+ Remember, our model for noise is a single normal distribution - so this value indicates how wide/variable this distribution is.
+ The model for noise implies points tend to be near the line; less likely to be far away (i.e. because the residuals are assumed to have a normal distribution with a mean of zero).
+ However, this value is not the uncertainty on any given prediction from the model, see Chapter \@ref(prediction) for details on that. 

### Introduction to the matrix form

There is another way to consider the simple linear regression model which allows for efficient notation and reflects how the best fit line is fitted in practice, as well as allowing computation of more complicated models in the future. The generic equation of the line can be given as:

$$ Y_i = \beta_0+\beta_1 X_i + \epsilon_i $$

where $Y_i$ is the $i$th $Y$ variable and $X_i$ is the $i$th predictor and $\epsilon_i$ the error associated with the $i$th point. 

For each datum in turn this would be

$$ Y_1 = \beta_0+\beta_1 X_1 + \epsilon_1 $$
$$ Y_2 = \beta_0+\beta_1 X_2 + \epsilon_2 $$
$$ Y_3 = \beta_0+\beta_1 X_3 + \epsilon_3 $$

etc.

The $Y$'s can be considered as single vector 

$$
\mathbf{Y} = \left[\begin{array}
{r}
Y_1  \\
Y_2  \\
\vdots\\
Y_n
\end{array}\right]
$$

Likewise the right-hand side of the equation can be broken up as


$$
\left[\begin{array}
{r}
\beta_0 + \beta_1 X_1 \\
 \beta_0 + \beta_1 X_2 \\  
\vdots \\
\beta_0+\beta_1 X_n
\end{array}\right] +
\left[\begin{array}
{r}
\epsilon_1 \\
 \epsilon_2 \\  
\vdots\\
\epsilon_n
\end{array}\right]
$$



which can be turned into (for those familiar with matrices and matrix multiplication) 


$$
\left[\begin{array}
{rr}
1 & X_1 \\
1 & X_2 \\  
\vdots & \vdots\\
1 & X_n
\end{array}\right] 
\left[\begin{array}
{r}
\beta_0 \\
\beta_1 \\  
\end{array}\right] +
\left[\begin{array}
{r}
\epsilon_1 \\
\epsilon_2 \\  
\vdots\\
\epsilon_n
\end{array}\right]
$$

If

$$
\mathbf{X} = 
\left[\begin{array}
{rr}
1 & X_1 \\
1 & X_2 \\  
\vdots & \vdots\\
1 & X_n
\end{array}\right] 
$$


$$
\mathbf{\beta} = 
\left[\begin{array}
{r}
\beta_0 \\
 \beta_1 \\  
\end{array}\right]
$$

and

$$
\mathbf{\epsilon} = 
\left[\begin{array}
{r}
\epsilon_1 \\
 \epsilon_2 \\  
\vdots \\
\epsilon_n
\end{array}\right]
$$





Then we get

$$ \mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}$$

Which for our fitted regression would be:

$$ \hat{\mathbf{y}} = \mathbf{X} \hat{\boldsymbol{\beta}} $$

Which is simply $y_i = \beta_0 + \beta_1 x_i$, for all values of $i$ in an economic way which allows scope for future complexity. 

To illustrate matrix calculations, we fit a simple regression to eight observations:

```{r}
  set.seed(345)
  x <- runif(8)
  y <- 2 + x*2 + rnorm(8)
  
  smallLM <- lm(y ~ x)

  plot(x, y)
  abline(coef(smallLM))
  modelEst <- as.vector(coef(smallLM))
  modelEst
```

The matrix *$X$* would be

```{r}
XMat <- cbind(rep(1, 8), x)
XMat
```

So if we (matrix) multiply this by the model coefficients we get our predicted values ($\hat{y}$). We can see this if we compare the results of the matrix multiplication `%*%` to the fitted values found from the regression model object. 

Compare: 
```{r}
 XMat %*% modelEst
```
to the fitted values:
```{r}
fitted(smallLM)
```

### Regression in practise

We can now return to the EIA example and use `DistCoast` to explain/predict `depth`:

+  `DistCoast` is the explanatory variable ($x_i$=`DistCoast`)
+  `depth` is the response variable ($y_i$=`depth`)
+  We have 5884 observations ($i=1,...,5884$) (observations right on the coast line have been removed)
+  However, we have `depth` values when `Distcoast` ($x$) is close to zero, but this is not guaranteed in many other situations; __it is ill-advised__ to assume a linear relationship holds outside the range of the observed data.

In this example, the slope is the change in `depth` (in m) for a 1 m increase in distance from the shore:

+  A slope $>0$ indicates a positive/increasing relationship
+  A slope$=0$ indicates no relationship (horizontal line)
+  A slope$<0$ indicates a negative/decreasing relationship
+  `depth` is measured on a continuous scale

We can start by modelling the differences between the data and the model using a normal distribution.

*N.B.*
+ There is not really a linear response in this example.
+ The relationship looks quite complex so maybe something else is going on. 
+ As we shall see, assuming a simple linear model (i.e. a straight line relationship) in this example might be inappropriate.  

#### Doing this in R

The filtered data have been stored in an object called `workingData`. The function used to fit a linear model is `lm`. 

```{r, echo=T, eval=T}
# Fit a regression using lm (linear model)
depthModel <- lm(depth ~ DistCoast,  data=workingData)
```

Note that when we write the regression equation we want to fit, we just need to specify $\textrm{response} \sim \textrm{explanatory variable}$. The intercept and gradient terms get included automatically.

To look at the output, it is useful to use `summary` as a wrapper function:

```{r, echo=T, eval=T}
summary(depthModel)
```

From the output we can obtain the regression coefficients:

+  the intercept: $\hat{\beta_0}=2.298$ 
+  the slope of the line: $\hat{\beta_1}= 1.244e^{-06}$ 
+  the estimated standard deviation of the errors: $s = \hat{\sigma}=4.32$ 

Thus, the fitted line (shown in Figure \@ref(fig:fitteddepth)) is:

$$\hat{\textrm{Depth}} = 2.298 + 1.244e^{-06}\textrm{DistCoast}$$


```{r fitteddepth, echo=FALSE, eval=TRUE, fig.cap='A scatterplot of depth and distance from coast with the least squares best fit line'}
p<- ggplot(x=DistCoast, y=depth, data=workingData) +
  geom_point(aes(DistCoast, depth), color='blue', alpha=0.3)+
  xlab('Distance from coast (m)')+
  ylab('Depth (m)')+ geom_abline(intercept=2.298,slope=1.244*10^-6)
p

```

It is worth looking at the summary output again, as well as an anova table output (equivalent to a one-way analysis of variance table). Using these functions in R tell you different information about the fitted model.

```{r, echo=F, eval=T}
# Fit a regression using lm (linear model)
depthModel <- lm(depth ~ DistCoast,  data=workingData)
```

The output from the `summary` function tells you what the regression coefficients are and whether they are significantly different from zero. 

```{r, echo=T, eval=T}
# Summary of fitted model
summary(depthModel)
```

The `anova` table tells you about the variation about the best fit line.

```{r, echo=T, eval=T}
# Analysis of variance table
anova(depthModel)
```

There are connections between the two tables. For example, the $t$ statistic (`t value`) in the `summary` statement, for `DistCoast` is directly related to the $F$ statistic in the `anova` table. 

$$t^2 = F$$

$$114.5^2 = 13112$$

We will return to the `anova` table later. 


## Summary {#SUMcorreg}

When we have two continuous variables, we are oftn interested in whether there is a relationship between them. A correlation coefficient measures the strength of a linear relationship. A simple linear regression describes the linear relationship when we want to use one variable to explain the other variable. Regression is a useful tool in statistics and can be extended to include many explanatory variables but first we consider the general framework of the linear model.    

### Learning outcomes 

At the end of this chapter you should understand

1. correlation and its constraints 
2. the need to be cautious in assigning causation, and
3. simple linear regression.

## Answers {#ANScorreg}

**Q13.1** Using the $r= \frac{Covar(XY)}{\sqrt(Var(x))\sqrt(Var(Y))}$ formula.

$$r= \frac{-2.5}{\sqrt(2.5)\sqrt(4)} = -0.791$$
**Q13.2** If the variance of $Z$ is calculated ignoring the covariance of $X$ and $Y$, then it will be overestimated as $X$ and $Y$ are being treated as independent (see earlier chapters for the addition rules for variances). 

**Q13.3** Assuming it is a real effect, there may be other reasons why _Booster_ consumers lose more weight than breakfast skippers. They may have a more healthy, or active, lifestyle more generally, for example.    

<!-- # Bibliography -->
