# Interactions and the Linear Model {#interac}

*To bring this back to a non-metaphorical level, I am suggesting that Black women can experience discrimination in ways that are both similar to and different from those experienced by
white women and Black men. Black women sometimes experience discrimination in ways similar to white women's experiences; sometimes they share very similar experiences with Black men. Yet
often they experience double-discrimination-the combined effects of practices which discriminate on the basis of race, and on the basis of sex. And sometimes, they experience discrimination as
Black women-not the sum of race and sex discrimination, but as Black women.*
<br/>Kimberlé Crenshaw [ -@Crenshaw1989] 


## Introduction {#INTinterac}

So far, we have only considered linear models that include __main effects__; so in the following model, 

$$y = \beta_0 + \beta_1x_1 + \beta_2 x_2 + \epsilon$$
where

+ $y$ is the response, 
+ $x_1$ and $x_2$ are continuous variable main effects; 
+ $\beta_0$ is the intercept, 
+ $\beta_1$ and $\beta_2$ are the gradients associated with $x_1$ and $x_2$, respectively, and 
+ $\epsilon$ is the error. 

Alternatively, we might have a categorical and continuous variable as main effects:
$$y = \beta_0 + \beta_{1_j} + \beta_2 x_2 + \epsilon$$
where 

+ $\beta_{1j}$ represents the intercept associated with a categorical variable level $j$ (note, $\beta_{1_1}=0$ if level 1 is used as a baseline/reference level). 

However, we might also be interested in including interactions between variables.

One question that might be asked of the environmental impact assessment (EIA) data, for example, is whether there is any evidence for a spatial re-distribution of bird density across construction phases (A,B or C) of the wind farm. We can do this by asking if a particular sort of density pattern in the `X` or `Y` spatial direction differs across phases, i.e. does the effect on density of `X` or `Y` differ between the levels of `phase`.  We call this sort of effect an **interaction**. 

+ Interaction is equivalent to the idea of "synergy" in chemistry or "intersectionality" in the humanities, or social sciences (see the quote by Kimberlé Crenshaw  at the beginning of the chapter).

__Example__ Say Drug A raised heartbeat by say 10 beats a minute and Drug B raises heartbeat by 20 beats per minute. Taken together they do not increase heartbeat by 30 beats per minute but reduce it by 10. There is a **non-additive** effect. 

But what does an interaction mean in statistical terms? In this chapter we illustrate statistical interaction and how to include them in linear models. 

### Fitting different models

Let's investigate human height as a function of leg length. 

```{r, echo=F, eval=T, suppress=T}
library (car)
set.seed (101)
LLL <- rnorm (100,110,4)
RLL <- rnorm(100, LLL,0.05)

TotalHeight <- (LLL+RLL)/2*1.8/1.1 +rnorm (100,0,4)

heightdata <- data.frame (TotalHeight, LLL, RLL)
heightdata <- rbind (heightdata, heightdata)
heightdata$TotalHeight[101:200] <- (LLL+RLL)/2*1.8/1.1 +rnorm (100,0,4)-10
heightdata$Sex <- c(rep ("M", 100), rep ("F", 100))
```

The data contains measurements for 100 males and 100 females, where the two groups differ in height (Figure \@ref(fig:boxplotheight)). We see that height differs by sex, but how is that affected by leg length?
 
```{r boxplotheight, echo=F, eval=T, fig.cap= "Boxplot of height by sex."}
boxplot (TotalHeight~Sex, data=heightdata, ylab= "Total Height (cm)")
```

\newpage

Exploratory analysis suggests that whilst females are generally shorter, the relationship between leg length and total height is the same in both sexes (Figure \@ref(fig:spheight)). There is a difference in the **intercept** but not in the **gradient** (slope) between the two sexes. 

```{r spheight, echo=F, eval=T, fig.cap="Scatterplot of left leg length and total height by sex."}
heightdata$SexNumeric <- ifelse (heightdata$Sex=="M",1,2)
          
plot (heightdata$LLL, heightdata$TotalHeight, col=heightdata$SexNumeric,
      pch=heightdata$SexNumeric, xlab="LLL (cm)", ylab= "Total Height (cm)")
legend(101, 192, legend=c("M","F"), col=1:2, pch=1:2)
```

This is confirmed by the regression analysis.

```{r, echo=T, eval=T}
# Fit model
model_height1 <- lm(TotalHeight ~ Sex + LLL, data=heightdata)
# ANOVA
anova(model_height1)
```

\newpage
```{r}
# Summary
summary (model_height1)
```
We can now obtain the best fit lines. The general model is:

$$\mathrm{TotalHeight} = \beta_0 + \beta_1\mathrm{Sex} + \beta_2\mathrm{LLL} + \epsilon$$
The best fit line for females (baseline) is: 
$$\mathrm{\widehat{TotalHeight}_{Females}} = -0.6919 + 1.5507 \times \mathrm{LLL}$$

The best fit line for males is: 
\begin{align}
\mathrm{\widehat{TotalHeight}_{Males}}& = -0.6919+ 10.1102 + 1.5507 \times \mathrm{LLL}\\
& = 9.4183 + 1.5507 \times \mathrm{LLL}\\
\end{align}

These lines can then be added to the scatter plots using the `abline` command where we supply the intercept (`a=`) and gradient (`b=`) (Figure \@ref(fig:spheight2)). 

\newpage

```{r spheight2, echo=T, fig.cap= "Scatterplot of left leg length and total height by sex with best fit lines with different intercept terms."}
# Create new variable for plotting colours/symbols
heightdata$SexNumeric <- ifelse(test=heightdata$Sex=="M", yes=1, no=2)
 
plot (heightdata$LLL, heightdata$TotalHeight, col=heightdata$SexNumeric, 
      pch=heightdata$SexNumeric, xlab="LLL (cm)", ylab= "Total Height (cm)")
# Add legend
legend(101, 192, legend=c("M","F"), col=1:2, pch=1:2)
# Red line for females
abline (a=-0.6919, b=1.5507, col=2)
# Black line for males
abline (a=(-0.6919 + 10.1102), b=1.5507)
```

Has this achieved what we want? We have not really allowed the model to consider the possibility that the relationship between LLL and total height differs between the sexes. We look at this now. 

## Fitting interaction terms

To test whether the relationship is different between the sexes we could fit regression lines to each sex independently and then examine the gradient terms in each model. However, this would halve our sample size in each case. An alternative is to allow the sex of the measured human to influence both intercept and the gradient of the best fit lines. Using `Sex` as a main effect allows for a different intercept (we have done this already). For an additional influence on gradient we need to specify an **interaction** term. We can do this in R by using the term `:` in the model formula between the variables of interest e.g.  

```{r, echo=T, eval=T}
# Fit model with interaction
model_height_interaction <- lm(TotalHeight ~ LLL + Sex + LLL:Sex,
                               data=heightdata)
summary(model_height_interaction)
```

Here the coefficient of the interaction term acts as a modifier of the gradient coefficient for the non-baseline sex (Figure \@ref(fig:plotheightint)). The fitted equations for each sex are given below. 

The best fit line for females (the baseline) is 
$$\mathrm{\widehat{TotalHeight}_{Females}} = 8.3556 + 1.4683 \times \mathrm{LLL}$$
The best fit line for males is 
\begin{align}
\mathrm{\widehat{TotalHeight}_{Males}} &= (8.3556 -7.9850) + (1.4683+0.1647) \times \mathrm{LLL}\\
& = 0.3706 + 1.633 \times \mathrm{LLL}\\
\end{align}

\newpage

```{r plotheightint, echo=T, eval=T, fig.cap= "Scatterplot of left leg length and total height by sex with best fit lines from an interaction."}
plot (heightdata$LLL, heightdata$TotalHeight, col=heightdata$SexNumeric, 
      pch=heightdata$SexNumeric, xlab="LLL (cm)", ylab= "Total Height (cm)")
legend(101, 192, legend=c("M","F"), col=1:2, pch=1:2)
# Red line for females
abline (a=8.3556, b=1.4683, col=2)
# Black line for males
abline (a=0.3706, b=1.633)
```

\newpage
Even allowing for a change in gradient, the best fit lines are not too dissimilar and an `anova` table reveals no significant interaction effect (this could also be inferred from the `summary` table).

```{r, echo=T, eval=T}
anova(model_height_interaction)
```

However, here is a very different sample of some rather tall male and female humans (perhaps they are Dutch!) (Figure \@ref(fig:tallplot)).

```{r tallplot, echo=F, eval=T, fig.cap= "Scatterplot of total height in leg length in tall humans."}
set.seed (101)
LLL <- rnorm (100,110,4)
RLL <- rnorm(100, LLL,0.05)

TotalHeight <- (LLL+RLL)/1.2+rnorm (100,0,4)

heightdata2 <- data.frame (TotalHeight, LLL, RLL)
heightdata2 <- rbind (heightdata2, heightdata2)
heightdata2$TotalHeight[101:200] <- (LLL+RLL)/1.9 +rnorm (100,0,4)+60
heightdata2$Sex <- c(rep ("M", 100), rep ("F", 100))
heightdata2$SexNumeric <- ifelse (heightdata2$Sex=="M",1,2)
plot (heightdata2$LLL, heightdata2$TotalHeight, col=as.numeric (heightdata2$SexNumeric), 
      pch=as.numeric (heightdata2$SexNumeric), xlab="LLL (cm)", ylab= "Total Height (cm)")
legend(101, 192, legend=c("M","F"), col=1:2, pch=1:2)

```

The relationship between `LLL` and `TotalHeight` is clearly different for each sex. This is confirmed by the `summary` output where the interaction term is significant.   

```{r, echo=T, eval=T}
model_height_interaction2 <- lm(TotalHeight ~ LLL + Sex + LLL:Sex, data=heightdata2)
anova (model_height_interaction2)
summary (model_height_interaction2)
```

The best fit lines are now:

For females (as females are the baseline):
$$\mathrm{\widehat{TotalHeight}_{Females}} = 78.4027 + 0.8842 \times \mathrm{LLL}$$
For males:
\begin{align}
\mathrm{\widehat{TotalHeight}_{Males}} & = (78.4027-78.0345) + (0.8842+0.7792) \times \mathrm{LLL}\\
& = 0.3682 + 1.6634 \times \mathrm{LLL}\\
\end{align}

which can then be plotted (Figure \@ref(fig:tallplot2)). 

```{r tallplot2, echo=T, eval=T, fig.cap= "Scatterplot of total height on left leg length in tall humans."} 
plot (heightdata2$LLL, heightdata2$TotalHeight, col=heightdata2$SexNumeric, 
      pch=heightdata2$SexNumeric, xlab="LLL (cm)", ylab= "Total Height (cm)")
legend(101, 192, legend=c("M","F"), col=1:2, pch=1:2)
# Red line for females
abline(a=78.4027, b=0.8842, col=2 )
# Black line for males
abline(a=0.3682, b=1.6634)
```

Two things to note:

+ the intercepts are *NOT* the same for each sex because the $y$ axis is not at $x = 0$. 

+ we have considered an interaction of a categorical variable with a continuous variable (= effect on the gradient).

### Specifying interactions in model formulae

If we were to write the above model as a general equation it would be:

$$\mathrm{TotalHeight} = \beta_0 + \beta_{\textrm{Sex}_j} + \beta_2\mathrm{LLL} + \gamma_{\textrm{Sex}_j}\mathrm{LLL} + \epsilon $$
or
$$\mathrm{TotalHeight} = \beta_0 + \beta_{\textrm{Sex}_j} + (\beta_2 + \gamma_{\textrm{Sex}_j})\mathrm{LLL} + \epsilon $$

where 

+ $\beta_0$ is the intercept,

+ $\beta_{\textrm{Sex}_j}$ is the intercept coefficient associated with Sex category $j$ (i.e. Male or Female),

+ $\beta_2$ is the gradient coefficient associated with $LLL$,

+ $\gamma_{\textrm{Sex}_j}$ is the gradient coefficient associated with Sex category $j$ for $LLL$, and 

+ $\epsilon$ is the error term.

+ In this example, $\beta_{\textrm{Sex}_\textrm{Female}}=0$ and $\gamma_{\textrm{Sex}_\textrm{Female}}=0$ because 'Female' is the baseline or reference level.

Writing this type of equation in even more general notation (e.g. with $y$ and $x$), then we have:

$$y = \beta_0 + \beta_{1_j} + \beta_2 x_2 + \gamma_{1_j}x_2 + \epsilon$$
$$y = \beta_0 + \beta_{1_j} + (\beta_2 + \gamma_{1_j})x_2 + \epsilon$$

where 

+ $\beta_0$ is the intercept,

+ $\beta_{1j}$ in the intercept coefficient associated with categorical variable 1 level $j$,

+ $\beta_2$ is gradient associated with continuous variable $x_2$,

+ $\gamma_{1j}$ is the gradient associated with categorical variable 1 level $j$ and $x_2$, and

+ $\epsilon$ is the error term.

An interaction of a continuous variable with another continuous variable would affect the gradient too, but in 3 dimensions (i.e. that interaction would have to be visualized using 3D plots or similar, e.g. section 3.6.4 and 3.6.5).

Algebraically, an interaction between two continuous variables ($x_1$ and $x_2$) is (simply) given as:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1x_2 + \epsilon$$

An interaction of two categorical variables would affect the intercepts and be illustrated (assuming no other variables) by a 3D bar chart or similar. 

Algebraically, an interaction between two categorical variables (denoted by $1$ and $2$) is given as:

$$y = \beta_0 + \beta_{1_j} + \beta_{2_k} + \gamma_{12_{jk}} + \epsilon$$
where 

+ $\beta_1$ and $\beta_2$ are the intercepts associated with categorical variables $1$ and $2$, level $j$ and $k$, respectively. 

+ $\gamma_{12_{jk}}$ is the intercept associated with the interaction between categorical variables 1 and 2, level $j$ and $k$. 

We can actually also have interactions of 3, or more, variables but the interpretation of such models can be very difficult and the data need to be well supported (i.e. you need cases of all the different combinations of the variable levels in the factorial case).

## Interactions in practise

In this section, we return to two data sets that we have introduced in previous chapters and look at including interactions. 

### EIA data

```{r, eval=T, echo=F}
wfdata<- read.csv("data/EIA_dataset_for_lectures_V1.0.csv", header=T)
```
We can implement an interaction(s) in our model to explain density by including `phase:X` and `phase:Y` terms:

\scriptsize

```{r, eval=F, echo=TRUE}

lm(formula = Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase 
   + X:phase + Y:phase, data=wfdata)

Coefficients:
                    Estimate Std. Error t value Pr(>|t|)    
(Intercept)        2.547e+03  3.097e+02   8.224  < 2e-16 ***
X                  6.906e-05  1.925e-05   3.588 0.000333 ***
Y                 -4.275e-04  5.080e-05  -8.416  < 2e-16 ***
DistCoast         -4.513e-07  1.343e-07  -3.360 0.000781 ***
depth             -5.288e-01  3.496e-02 -15.125  < 2e-16 ***
as.factor(month)2  4.771e-01  5.363e-01   0.890 0.373617    
as.factor(month)3  3.116e+00  4.624e-01   6.737 1.64e-11 ***
as.factor(month)4  6.198e-01  4.587e-01   1.351 0.176611    
phaseB             1.068e+02  3.254e+02   0.328 0.742716    
phaseC             1.431e+03  6.516e+02   2.196 0.028084 *  
X:phaseB           7.012e-05  2.572e-05   2.726 0.006407 ** 
X:phaseC           2.532e-05  3.183e-05   0.796 0.426314    
Y:phaseB          -2.559e-05  5.329e-05  -0.480 0.631115    
Y:phaseC          -2.392e-04  1.078e-04  -2.218 0.026566 *  
---
Signif. codes:  0 *** 0.001 ** 0.01 * 0.05 . 0.1   1

Residual standard error: 27.85 on 31488 degrees of freedom
Multiple R-squared:  0.01483,	Adjusted R-squared:  0.01442 
F-statistic: 36.46 on 13 and 31488 DF,  p-value: < 2.2e-16

```

\normalsize

There are rules for including interaction terms in a model:

+ (Unless you are an advanced user) if you have an interaction term you should also include the main effects terms associated with the interaction. In the example above, the model includes `phase`, `X` and `Y` as main effects. 

+ If the interaction is significant, the $p$-values associated with the main effects are irrelevant and so the main effects are retained. 

+ If $p$-value selection is in operation and the interaction is removed, the main effects should not be removed before re-evaluating the model. 

+ Interactions always come last in the sequence of predictors. 

__A phase-based interaction term__

In our new interaction-based model (above) we have:

\begin{equation}
y_{it}=\beta_{0}+\beta_1x_{1i}+\beta_2x_{2i}+\beta_3x_{3i}+...+\beta_{13}x_{13i}
\end{equation}

where $\beta_{1}-\beta_9$ and $x_{1i}-x_{9i}$ are as described before and relate to `X`,
`Y`, `DistCoast`, `depth`,  `month` and `phase`. The new aspects of the output are as follows:

+  `X:phaseB`: $\beta_{10}$ is the expected change in the slope coefficient for the X relationship in phase B compared with the X relationship in phase A
+  `X:phaseC`: $\beta_{11}$ is the expected change in the slope coefficient for the X relationship in phase C compared with the X relationship in phase A
+  `Y:phaseB`: $\beta_{12}$ is the expected change in the slope coefficient for the Y relationship in phase B compared with the Y relationship in phase A
+  `Y:phaseC`: $\beta_{13}$ is the expected change in the slope coefficient for the Y relationship in phase C compared with the Y relationship in  phase A

The uncertainty associated with the interaction-based estimates result in:

+ no statistically significant difference between the `X`-slope coefficient for phase A compared with phase C (`X:phaseC`; $p$-value=0.426)  
+  no significant difference between the `Y`-slope coefficient for phase B compared with phase A (`Y:phaseB`; $p$-value=0.631). 

Looking at the ANOVA table below, (overall) there is evidence for a `X`-`phase` interaction ($p$-value=0.0216) but no evidence for a `Y`-`phase` interaction term ($p$-value=0.0848).

```{r, eval=F}
Anova Table (Type II tests)

Response: Density
                   Sum Sq    Df  F value    Pr(>F)    
X                   65350     1  84.2826 < 2.2e-16 ***
Y                  111907     1 144.3281 < 2.2e-16 ***
DistCoast            8753     1  11.2886 0.0007808 ***
depth              177386     1 228.7771 < 2.2e-16 ***
as.factor(month)    49242     3  21.1693 1.079e-13 ***
phase                 374     2   0.2410 0.7858532    
X:phase              5946     2   3.8342 0.0216287 *  
Y:phase              3827     2   2.4681 0.0847604 .  
Residuals        24414670 31488                       

```

+ If we remove the `Y`-`phase` interaction from the model then all terms are now significant in the model.
+ Note, while the `phase` term considered alone is not significant in the model, it forms part of the interaction term and so is typically retained in the model regardless.
+ There are not grounds to reduce this model further, if backwards selection was being undertaken. 

### Medical data

```{r, echo=F}
meddata<- read.csv("data/medical_dataset_for_R_lectures_v1.0.csv")
```

```{r echo=FALSE}
# NOT SHOWN TO STUDENTS
meddata<- meddata[which(is.na(meddata$vitdresul)==FALSE),]
meddata<- meddata[which(meddata$folate<200),]
# removing the irritating whitespaces in the character columns;
meddata$vitdc<-as.factor(gsub(" ", "", meddata$vitdc, fixed = TRUE))
meddata$vitbc<-as.factor(gsub(" ", "", meddata$vitbc, fixed = TRUE))
meddata$teonpres<-ifelse(test = meddata$TEON=='Yes', 1, 0)
```

A previous model considered the influence of `TEON`, `folate` and `gender` on vitamin D level separately. But what if we believed `folate` potentially affected vitamin D level but in a different way depending on `gender`? This can be investigated using an interaction term:

```{r, echo = T}
multiReg_lm <- lm(vitdresul ~ TEON + folate + gend + folate:gend, data=meddata)
summary(multiReg_lm)
anova(multiReg_lm)
```
In this case, there is less evidence to support the existence of an interaction effect. 

The syntax for an equation in R can be shortened: `folate + gend + folate:gend` can be abbreviated by `folate*gend`. So the following commands are identical to those above. 

```{r, echo = T}
multiReg_lm <- lm(vitdresul ~ TEON + folate * gend, data=meddata)
anova(multiReg_lm)
```

It is also worth considering what happens if the command `Anova` is used rather than `anova` above. 

```{r, echo = T}
multiReg_lm <- lm(vitdresul ~ TEON + folate * gend + folate, data=meddata)
Anova(multiReg_lm)
```

In this case, the $p$-value associated with the last term is the same as we might expect. But the second to last term is also the same, which may come as a surprise. `Anova` has to follow the *principle of marginality*, the interaction must come last. The main effects are now given as if they were second from last. Thus `gend` remained the same (it was second from last before) but the $p$-value associated with `TEON` and `folate` are different. Notice the `Anova` table is subtitled "Type II tests". Remember ordinary `anova` is Type I, the sum of squares are considered sequentially. Type II sums of squares are sum of squares where the term is considered as if it was last in the model.   

## Model selection and interactions

If backwards model selection is being undertaken, interaction terms should be considered first for removal (following the principle of marginality) and then other main effects not associated with the interaction as before. Main effects associated with the interaction should not be considered (for rejection) unless the interaction is rejected. 

If forwards selection is being undertaken, main effects should be added first and then interactions added. 

### Backwards selection in the EIA data set 

Backwards selection is illustrated using different selection criterion for the following model. 

```{r, eval=T, echo=F}
wfdata$Density <- wfdata$response/wfdata$area
```

```{r}
modelcomplex <- lm(Density ~ X + Y + DistCoast + depth + as.factor(month) +  
                     phase + X:phase + Y:phase, data=wfdata)
```

#### Using $p$-values

Using `Anova` (Type II sum of squares) we can simultaneously consider both interactions at the same time (as we saw above).  So in the first "round" we consider `X:phase` and `Y:phase` interactions only. 

```{r, eval=T, echo=F}
Anova(modelcomplex)
```

If we are using a significance level ($\alpha$) of 0.05, `Y:phase` has the highest $p$-value and it is >0.05 so we reject the `Y:phase` term and refit. 

```{r}
modelcomplex2 <- lm(Density ~ X + Y + DistCoast + depth + as.factor(month)  
                    + phase + X:phase, data=wfdata)
Anova(modelcomplex2)
```

All other terms are significant except for `phase` but that is covered by the interaction term so is retained, so we stop there. Hence, the best fit model using p values is:

`Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase + X:phase`.

#### Using AIC

In this case, we have to consider the AIC of the original model above and the AICs of the model without the `X:phase` term and then the starting model without the `Y:phase` term.

\small
```{r, eval=T, echo=T}
AIC(lm(Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase + X:phase + Y:phase, 
       data=wfdata))
AIC(lm(Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase + X:phase, data=wfdata))
AIC(lm(Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase + Y:phase, data=wfdata))

```
\normalsize

The AICs are 299008.1,  299009 and  299011.8, respectively. 

OK, we should now consider the AICs of the main effects (except `X`, `Y` and `phase`) and see if any of these are less than 299008.1. 

\small
```{r, eval=T, echo=T}
AIC (lm(Density ~ X + Y + depth + as.factor(month) + X:phase + Y:phase, data=wfdata))
AIC (lm(Density ~ X + Y + DistCoast + as.factor(month) + phase + X:phase + Y:phase, data=wfdata))
AIC (lm(Density ~ X + Y + DistCoast + depth + phase + X:phase+Y:phase, data=wfdata))
```
\normalsize

The AICs are 299014.4, 299234.1 and  299065.6, so we have no grounds to reduce the model further. So our best model using AIC is:

`Density ~ X + Y + DistCoast + depth + as.factor(month) +  phase + X:phase+Y:phase`. 

One thing to note from this analysis:

+ **$p$-value model selection and AIC (or similar score) selection do not necessarily produce the same answers!**

### Backwards selection in the medical data set

Let's see what happens with the medical data set.

#### Using $p$-values

In this case, the model called `multiReg_lm` indicated the `folate:gend` term was associated with a $p$-value>0.05, so we remove this term and then consider the $p$-values associated with all of the main effects. Again, the quickest way to do this is to use `Anova` rather than consider lots of sequential models using `anova`. 
```{r, echo = T, eval=T}
multiReg_lm <- lm(vitdresul ~ TEON + folate + gend, data=meddata)
Anova(multiReg_lm)
```

The term `gend` has the highest $p$-value >0.05, so we reject that term and go again.  

```{r, echo = T, eval=T}
multiReg_lm <- lm(vitdresul ~ TEON + folate, data=meddata)
Anova(multiReg_lm)
```
Now we have grounds for rejecting `folate` and we are left with a model with `TEON` only. 

```{r, echo = T, eval=T}
multiReg_lm <- lm(vitdresul ~ TEON, data=meddata)
Anova(multiReg_lm)
summary(multiReg_lm)
```

This is the final chosen model. 

#### Using AIC

In this case, we first consider the AIC of a model with and without an interaction. 

```{r, echo = T, eval=T}
AIC(lm(vitdresul ~ TEON + folate + gend + folate:gend, data=meddata))
AIC(lm(vitdresul ~ TEON + folate + gend, data=meddata))

```

AIC is more "generous" than $p$-values so the interaction remains. We should now compare the model without TEON but with the interaction. 

```{r, echo = T, eval=T}
AIC(lm(vitdresul ~ folate + gend + folate:gend, data=meddata))

```

This AIC is not lower than 376.2377, so we stay with the model:

`vitdresul ~ TEON + folate + gend + folate:gend` model.

Note that this is a different model to that chosen using P-values. 

**Q16.1**	Dr Teuthis  was interested in predicting entire length (EL) of a giant squid from mantle length (ML). Mantle length is the length of just the body, or "mantle", and the entire length is the length of the body plus head plus tentacles. See Paxton[-@Paxton2016] for an actual analysis of the data. Dr Teuthis was also interested in whether the ratio of total length to mantle length is different by sex (Male, Female or Not Known, NK). The following analysis was performed. 
```{r, eval=F}
modelMLtoEL <- lm(EL ~ ML + Sex + ML:Sex, data=squidtemp)
```

Write down the general equation for the model using $\beta$s. 

**Q16.2**	Explain this model in words.  

**Q16.3**	How might you illustrate the data and the fitted model graphically?

**Q16.4**	Dr Teuthis used the following command to generate an ANOVA table. 


``` {R, eval=F}
Anova(modelMLtoEL)

Type II tests

Response: EL`

            Sum Sq   Df   F value    Pr(>F)       
  ML        179.634  1    74.4497   9.242e-11 ***   
  Sex         4.473  2     0.9270   0.8039          
  ML:Sex      0.871  2     0.1806   0.4354          
  Residuals  98.926 41                              
```

Based on this information what would be your next step in the modelling process?

**Q16.5**	Here is the summary of the above model object. 
```{R, eval=F}

Call:
lm(formula = EL ~ ML * Sex, data = squidtemp)

Residuals:
    Min      1Q  Median      3Q     Max 
-2.6997 -0.8099 -0.1017  0.6817  4.5252 

Coefficients:
                Estimate Std. Error t value Pr(>|t|)    
(Intercept)       1.9844     1.5836   1.253 0.217267    
ML                3.6121     1.0179   3.548 0.000988 ***
SexMale          -0.9284     2.3848  -0.389 0.699078    
SexNK            -0.8995     1.7909  -0.502 0.618185    
ML:SexMale        1.1386     1.8946   0.601 0.551175    
ML:SexNK          0.3224     1.1516   0.280 0.780925    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 1.553 on 41 degrees of freedom
Multiple R-squared:  0.663,     Adjusted R-squared:  0.6219 
F-statistic: 16.13 on 5 and 41 DF, p-value: 8.858e-09

```

Write down the fitted equations for female, male and sex not known (NK). 
Hence, estimate the entire length for a female, male and unknown sex squid with a mantle length of 2 metres.

## Summary {#SUMinterac}

Interactions are extremely useful but should be used carefully, always understanding what the interaction represents in terms of means and gradients. 

### Learning objectives
At the end of this chapter, you should understand how to: 

1. use interactions in understanding the relationships of data,
2. interpret R output containing interactions.

## Answers {#ANSinterac}

**Q16.1**
$$
EL=\ {\beta{}}_0+\ {\beta{}}_1ML+\ {\beta{}}_2Sex+\ {\beta{}}_3ML.Sex+\ \in{}
$$

**Q16.2**	The model describes the hypothesised relationship between mantle length and entire length, and how this relationship differs between the sex categories. 

**Q16.3**	It might be illustrated using a scatterplot with mantle length on the x-axis and entire length on the y-axis with the points different colours, or symbols, to represent the different sex categories. Three best-fit regression lines are estimated (for Males, Females and NK) and superimposed on to the plot, representing the relationship between EL and ML for each sex. 

**Q16.4** This model contains an interaction term and so the interaction term should be considered first (because of the principle of marginality) before individual terms in the interaction can be excluded as main effect terms. In this case, the interaction should be excluded because it is not significant ($p$-value=0.44). Note, if an interaction was not included in the model, then remove the variable with the highest probability if it is greater than the significance level used for testing (i.e. if the term is not significant). 

**Q16.5** Female is used as the reference level.
\raggedright
Females:\hspace{15pt} $\hat{EL}=1.9844+3.6121\times{}ML\ $
\raggedright
Males:\hspace{15pt}\hspace{15pt}
$\hat{EL}=1.9844-0.9284+\left(3.6121+1.1386\right)\times{}ML=1.056+4.7507\times{}ML\
$
\raggedright
Not known:\hspace{15pt}
$\hat{EL}=1.9844-0.8995+\left(3.6121+0.3224\right)\times{}ML=1.0849+3.9345\times{}ML\
${\raggedright
Hence  
Female:\hspace{15pt}\hspace{15pt}1.9844 + 3.6121 * 2 = 9.209 m  
\raggedright    
Male:\hspace{15pt}\hspace{15pt}1.056 + 4.7507 * 2 = 10.56 m  
\raggedright  
Not known:\hspace{15pt}1.0849 + 3.9345 * 2 = 8.95 m

<!-- # Bibliography -->
