--- 
title: "An Introduction to Statistics"
author: "Authors"
date: '`r format(Sys.Date(), "%d %B %Y")`'
site: bookdown::bookdown_site
documentclass: krantz
classoption: oneside
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
---


# Welcome {-}

```{r setup, echo=FALSE, warning=FALSE}
library(knitr)
## set knitr options
opts_chunk$set(cache = F, tidy.opts = list(width.cutoff = 55), tidy = F, 
               fig.align = "center", fig.width = 5, fig.height = 5, 
               multCode = F, renderTask = T, renderSol = T, 
               warning=FALSE, message=FALSE, dev='png', comment='')
```

```{r, child = "_setup.Rmd", include = F, purl = F, cache = F}
```

Welcome to an introduction to statistics.

This is an introductory course on statistical using the statistical programming environment `R`. The focus will be on how to undertake an exploratory data analysis in `R`. The target audience is anyone who wants to understand basic statistical concepts to summarise and visualise data and undertake simple analysis. The progression will be via further courses in semester 2.

## Prerequisites {-}

* anything here?

## Working through the module {-}

This module consists of a series of linked chapters. Most chapter consists of

1. Core material.
2. A computer practical to practice implementing in the program `R`, the methods learnt in the core text (except Chapter 1), and
3. A series of questions which contribute to the overall assessment for this module.

It is a good idea to complete all three sections before moving to the next chapter because the ideas and code build upon the previous chapter. (**they will need to complete all the assessment questions to complete the module. can make it such that they have to have tried the computer practical (get some kind of check mark against it?)**)

The core material consists of:

+ explanatory material containing the core factual content of the module,
+ exercise/reflective questions (**in bold**) with answers provided at the end of the chapter (**when in print. when in the online version will it be there and then?**),
+ additional explanations of specific ideas or concepts accessed by hovering/clicking on the link (**TBD. these may be videos or shinys**),
+ sections of `R` code to illustrate how to implement the ideas developed in each chapter - the code appears in `courier` font in a shaded box.

Common notation can be found in xxx (**an appendix?**) 

## Recommended reading {-}

<!-- I highly recommend the following books: -->

<!-- * [Statistics: An Introduction using R](https://www.wiley.com/en-gb/Statistics%3A+An+Introduction+using+R-p-9780470022986) -->

<!-- * [Linear models with R](https://www.crcpress.com/Linear-Models-with-R/Faraway/p/book/9781439887332) -->


## Data files {-}

<!-- All data files can be downloaded as a ZIP file from somewhere.... -->

## Acknowledgements {-}



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'ggplot2', 'Hmisc', 'epitools', 'ggplot2', 'datasauRus', 'MuMIn', 'effects', 'ggExtra', 'boot', 'simpleboot', 'car', 'onewaytests', 'pgirmess', 'dplyr', 'tidyr', 'car'
), 'packages.bib')
```
